---
title: "Differential Expression Dataset: Human Kidney"
output: 
  flexdashboard::flex_dashboard:
  orientation: columns
  vertical_layout: fill
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(shiny)

load("ShInput/ShinyObj.RData")
# translate p-vals to -log space
de_Results$Significance <- -log10(de_Results$FDR)
fullPath_df$Significance <- -log10(fullPath_df$padj)

# Colorset
clrs <- c("gray", "#515151", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00")
names(clrs) <- c("nosig", "nodelt", "normal", "DKD", "glomerulus", "tubule")
# Set colors based on cutoffs, break up Contrast
de_Results <- de_Results %>%
  dplyr::mutate(clr = 'nodelt') %>%
  tidyr::separate(Contrast, into = c('right','left'), sep = ' - ', remove = F)
de_Results$clr[de_Results$Estimate >= 0.5] <- de_Results$right[which(de_Results$Estimate >= 0.5)]
de_Results$clr[de_Results$Estimate <= -0.5] <- de_Results$left[which(de_Results$Estimate <= -0.5)]
de_Results$clr[de_Results$FDR > 0.05] <- 'nosig'

```


DE Results
======

Column {.sidebar}
------
### Graph Criteria

This is an interactive tool to investigate the differential Expression results from 10,048 genes over 221 kidney samples. Choose first the test of interest, then the subset of data. A final dropdown menu offers the top 20 enriched Reactome pathways, and the choice will highlight the genes in pink for that pathway.

``` {r}

selectInput(inputId = "Test", label = "Select Test", choices = unique(de_Results$Contrast),
            selected = 1, multiple = FALSE)

selectInput(inputId = "Subset", label = "Select Data Subset", 
            choices = NULL,
            selected = NULL, multiple = FALSE)

selectInput(inputId = "Pathways", label = "Select Pathway of Interest", 
            choices = NULL,
            selected = NULL, multiple = FALSE)

observeEvent(input$Test,
             {updateSelectInput(session, input = "Subset",
                                choices = unique(de_Results[de_Results$Contrast == input$Test,]$Subset))})
observeEvent(input$Subset,
             {updateSelectInput(session, input = "Pathways",
                                choices = sigpath_df[sigpath_df$Subset == input$Subset &
                                                       sigpath_df$Contrast == input$Test, ]$pathway)})

```

Review the data in a volcano plot, MA plot, or pathway volcano plot. On the Expression page, a simple violin/boxplot by data subset allows review of log2 mean counts by gene.

Column {.tabset}
-------------------

### Volcano Plot

Cutoffs set to +/- 0.5 log2 fold change and FDR < 0.05.

```{r}
de_rct <- reactive({
  de_Results %>%
    filter(Contrast == input$Test) %>%
    filter(Subset == input$Subset)})
# vector of genes for pathway of interest
genes <- reactive({
  gs <- sigpath_df$leadingEdge[sigpath_df$pathway == input$Pathways &
                                 sigpath_df$Subset == input$Subset &
                                 sigpath_df$Contrast == input$Test]
  str_split(gs, pattern = ',')[[1]]
})

renderPlotly({
  sub <- de_rct()
  xlabel <- str_split(str_to_title(input$Test), ' - ')[[1]]
  xlabel <- paste0("← Higher in ", xlabel[2], "  |  Higher in ", xlabel[1], " →")
  mx <- max(abs(sub$Estimate))
  ggplotly(ggplot(sub[!sub$Gene %in% genes(), ], 
                  aes(x = Estimate, y = Significance, color = clr, text = Gene)) +
    geom_vline(xintercept = 0.5, lty = 'dashed', lwd = 0.5, color = 'gray') +
    geom_vline(xintercept = -0.5, lty = 'dashed', lwd = 0.5, color = 'gray') +
    geom_hline(aes(yintercept = -log10(0.05)),
                 color = 'deeppink', alpha = 0.5, linetype = 'dashed', linewidth = 0.5) +
    geom_point(size = 2, alpha = 0.65) +
    scale_colour_manual(values = clrs) +
    geom_point(data = sub[sub$Gene %in% genes(), ], color = "hotpink") +
    theme(legend.position="none") +
    theme_bw() +
    xlim(-mx * 1.1, mx * 1.1) +
    labs(title = str_replace(str_to_title(input$Test), ' - ', ' versus '),
         subtitle = paste0("Subset: ", input$Subset),
         x = paste(xlabel, "\nlog2 Fold Change"),
         y = "-log10(FDR)"), #expression(paste(-log[10], italic(" (p)")))
    tooltip = "text") %>% layout(showlegend = FALSE)
})

```

### MA Plot

Colored points indicate genes beyond fold change and FDR cutoffs. Pink points are genes in the highlighted pathway.

```{r}
# Layer points atop one another for MA plot
renderPlotly({
  sub <- de_rct()
  xlabel <- str_split(str_to_title(input$Test), ' - ')[[1]]
  xlabel <- paste0("← Higher in ", xlabel[2], "  |  Higher in ", xlabel[1], " →")
  mx <- max(abs(sub$Estimate))
  ggplotly(ggplot(sub[!sub$Gene %in% genes() & sub$clr %in% c("nosig", "nodelt"), ], 
                  aes(x = mn_exp, y = Estimate, color = clr, text = Gene)) +
    geom_point(size = 2, alpha = 0.65) +
    scale_color_manual(values = clrs) +
    geom_point(data = sub[!sub$Gene %in% genes() & !sub$clr %in% c("nosig", "nodelt"), ]) +
    geom_point(data = sub[sub$Gene %in% genes(), ], color = "hotpink") +
    theme(legend.position="none") +
    theme_bw() +
    ylim(-mx * 1.1, mx * 1.1) +
    labs(title = str_replace(str_to_title(input$Test), ' - ', ' versus '),
         subtitle = paste0("Subset: ", input$Subset),
         y = paste(xlabel, "\nlog2 Fold Change"),
         x = "Mean log2 Expression"), 
    tooltip = "text") %>% layout(showlegend = FALSE)
})

```

### Pathway Analysis

```{r}
pw_rct <- reactive({
  fullPath_df %>%
    filter(Contrast == input$Test) %>%
    filter(Subset == input$Subset)})

renderPlotly({
  xlabel <- str_split(str_to_title(input$Test), ' - ')[[1]]
  xlabel <- paste0("← Enriched in ", xlabel[2], "  |  Enriched in ", xlabel[1], " →")
  mx <- max(abs(pw_rct()$NES))
  ggplotly(ggplot(pw_rct()[pw_rct()$pathway != input$Pathways, ], 
                  aes(x = NES, y = Significance, text = pathway)) +
    geom_point(color = "gray25") +
    geom_point(data = pw_rct()[pw_rct()$pathway == input$Pathways, ], color = "hotpink") +
    theme_bw() +
    xlim(-mx * 1.1, mx * 1.1) +
    labs(title = str_replace(str_to_title(input$Test), ' - ', ' versus '),
         y = "-log(Adjusted p-value)",
         x = paste(xlabel, "\nNormalized Enrichment Score")), 
  tooltip = "text") %>% layout(showlegend = FALSE)

})


```

Expression
=======

Column {.sidebar}
-------------------

### Mean Log2 Expression

Note that for this public dataset from the `GeoMxWorkflows` vignette, counts have been Q3 normalized and log2 transformed.

Column
-------------------

```{r}

renderPlotly({
  ggplotly(ggplot(de_Results, aes(x = Subset, y = mn_exp, fill = Subset)) +
    geom_boxplot() +
    scale_fill_manual(values = clrs) +
    geom_violin(alpha = 0.3) +
    # scale_color_manual(values = clrs) +
    labs(title = "Log2 Mean Expression by Data Subset", y = "Log2 Expression") +
    theme_bw() +
    theme(axis.title.x = element_blank(),
          legend.position="none"))
})

```


